#  Building Machine Learning Models {#sec-ml-in-practice}

Building machine learning models requires more than understanding the underlying mathematics. While theory helps avoid conceptual mistakes and guides the choice of methods, translating these ideas into effective, real-world models involves a series of practical decisions. These include how to preprocess and transform features, incorporate domain knowledge to select relevant predictors, manage computational constraints, and systematically tune and compare algorithms. In many cases, we also combine models through ensembling to improve predictive performance. 

In this chapter, we bring these elements together through a case study based on the MNIST handwritten-digit dataset. Using this example, we illustrate how the concepts developed in earlier chapters, particularly model evaluation, selection, and tuning, fit into a coherent machine learning workflow.

Before turning to the case study, we introduce the **caret** package, which provides a unified interface to a wide range of machine learning algorithms. By standardizing syntax and workflow, **caret** makes code more consistent, readable, and easier to maintain. Although we use **caret** throughout this chapter, similar frameworks exist in both R and Python, including **tidymodels** and **mlr3** in R, and **scikit-learn** and **PyTorch** in Python. Our goal is not to cover any one framework exhaustively, but to illustrate how such tools structure the modeling process so that you can readily adapt to others.

## The caret package {#sec-caret}

We have already learned about several machine learning algorithms. Many of these algorithms are implemented in R. However, they are distributed via different packages, developed by different authors, and often use different syntax. The __caret__ package tries to consolidate these differences and provide consistency. It currently includes over 200 different methods, which are summarized in the __caret__ package manual^[[https://topepo.github.io/caret/available-models.html](https://topepo.github.io/caret/available-models.html)]. Keep in mind that __caret__ does not include the packages needed to run each possible algorithm. To apply a machine learning method through __caret__, you still need to install the library that implements the method. The required packages for each method are described in the package manual.

To illustrate the __caret__ package and its key functions, we will first use the `mnist_27` dataset introduced in @sec-mnist that contains only images of 2s and 7s and two predictors. This smaller dataset allows us to run code quickly while focusing on learning the package. Once we are comfortable with the package, we will use the full MNIST dataset for realistic performance evaluation and demonstrate how the concepts scale to larger datasets.

### The `train` function 

The R functions that fit machine learning algorithms are all slightly different. Functions such as `lm`, `glm`, `qda`, `lda`, `knn3`, `rpart`, and `randomForest` use different syntax, have different argument names, and produce objects of different types.

The __caret__ `train` function lets us train different algorithms using similar syntax. So, for example, we can type the following to train three different models:

```{r}
#| message: false
#| warning: false
#| cache: false
library(dslabs)
library(caret)
```
```{r}
#| message: false
#| warning: false
train_glm <- train(y ~ ., method = "glm", data = mnist_27$train)
train_qda <- train(y ~ ., method = "qda", data = mnist_27$train)
train_knn <- train(y ~ ., method = "knn", data = mnist_27$train)
```

As we explain in more detail in @sec-caret-cv, the `train` function selects parameters for you using a resampling method to estimate the MSE, with bootstrap as the default.

### The `predict` function

The predict function is one of the most useful tools in R for machine learning and statistical modeling in general. It takes a fitted model object and a data frame containing the features $\mathbf{x}$ for which we want to make predictions, then returns the corresponding predicted values.

Here is an example with logistic regression:

```{r}
fit <- glm(y ~ ., family = "binomial", data = mnist_27$train)
p_hat <- predict(fit, newdata = mnist_27$test)
```

In this case, `predict` is simply computing 

$$
\hat{p}(\mathbf{x}) = g^{-1}\left(\hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 \right) \text{ with } g(p) = \log\frac{p}{1-p} \implies g^{-1}(\mu) = \frac{1}{1-e^{-\mu}}
$$

for the `x_1` and `x_2` in the test set `mnist_27$test`. With these estimates in place, we can make our predictions and compute our accuracy:

```{r}
y_hat <- factor(ifelse(p_hat > 0.5, 7, 2))
```

However, note that `predict` does not always return objects of the same type; it depends on what type of object it is applied to. To learn about the specifics, you need to look at the help file specific to the type of fit object that is being used. 

`predict` is actually a special type of function in R called a _generic function_. Generic functions call other functions depending on the kind of object they receive. So if `predict` receives an object coming out of the `lm` function, it will call `predict.lm`. If it receives an object coming out of `glm`, it calls `predict.glm`. If the fit is from `knn3`, it calls `predict.knn3`, and so on. These functions are similar but not identical. You can learn more about the differences by reading the help files:

```{r, eval = FALSE}
?predict.glm
?predict.qda
?predict.knn3
```

There are many other versions of `predict`, and many machine learning algorithms define their own version.

As with `train`, the **caret** package unifies the use of `predict` with the function `predict.train`. This function takes the output of `train` and produces predictions of categories using `type = "raw"` (default) or estimates of $p(\mathbf{x})$ using `type = "prob"`.

The code looks the same for all methods:

```{r}
y_hat_glm <- predict(train_glm, mnist_27$test, type = "raw")
y_hat_qda <- predict(train_qda, mnist_27$test, type = "raw")
y_hat_knn <- predict(train_knn, mnist_27$test, type = "raw")
```

This permits us to quickly compare the algorithms. For example, we can compare the accuracy like this:

```{r}
fits <- list(glm = train_glm, qda = train_qda, knn = train_knn)
sapply(fits, function(fit) {
  mean(predict(fit, mnist_27$test) == mnist_27$test$y)
})
```


### Resampling {#sec-caret-cv}

When an algorithm includes a tuning parameter, `train` automatically uses a resampling method to estimate MSE and decide among a few default candidate values. To find out what parameter or parameters are optimized, you can read the **caret** manual (see footnote 1) or study the output of: 

```{r, eval=FALSE}
modelLookup("knn")
```

To obtain all the details of how **caret** implements kNN, you can use:

```{r, eval=FALSE}
getModelInfo("knn")
```

If we run it with default values: 

```{r}
train_knn <- train(y ~ ., method = "knn", data = mnist_27$train)
```

you can quickly see the estimated MSEs using the `ggplot` function. The argument `highlight` highlights the max:

```{r caret-highlight}
#| warning: false
ggplot(train_knn, highlight = TRUE)
```

We see that the accuracy is estimated to be around 0.81, which is not much higher than the 0.775 we obtained with logistic regression and QDA. However, for the `knn` method, the default is to try $k=5,7,9$, and it is possible that we can do better by trying out more values. We change this using the `tuneGrid` argument. The grid of values must be supplied by a data frame with the parameter names as specified in the `modelLookup` output.  Below, we present an example where we try out values between 1 and 149. To do this with __caret__, we need to define a column named `k`, so we use this:  `data.frame(k = seq(1, 149, 2))`. 

:::{.callout-note}
### Limitations of Bootstrapping for kNN

By default, the **caret** `train` function uses bootstrap resampling. However, bootstrapping can be problematic when estimating mean squared error (MSE) for k-nearest neighbors (kNN), particularly for small values of $k$.

Because bootstrap samples are drawn with replacement, the same observation can appear multiple times in a training set. For kNN, this can artificially inflate the influence of duplicated points, since identical observations may be treated as distinct neighbors. As a result, the model can achieve unrealistically low error estimates, leading to overly optimistic performance.

To avoid this issue, we use cross-validation instead. For example, we can specify:

```{r}
cv20_ctrl <- trainControl(method = "cv", number = 20)
```

which performs 20-fold cross-validation, using 95% of the data for training in each split.
:::

Here is an example of running `train` with cross-validation and a custom grid of $k$ values. Note that when running this code, we are fitting 75 versions of kNN to 25 bootstrapped samples. Since we are fitting $75 \times 25$ = 1,875 kNN models, running this code might take several seconds.

```{r train-knn-plot}
#| warning: false
set.seed(2001)
train_knn <- train(y ~ ., method = "knn", data = mnist_27$train,
  tuneGrid = data.frame(k = seq(1, 149, 2)),
  trControl = cv20_ctrl
)
ggplot(train_knn, highlight = TRUE)
```

Resampling methods involve random sampling steps, so repeated runs of the same code can produce different outcomes. To ensure results can be reproduced exactly, we set the random seed at the start of the analysis. 

The `results` component of the `train` output includes several summary statistics related to the variability of the cross-validation estimates:

```{r}
names(train_knn$results)
```
You can learn many more details about the **caret** package from the manual (see footnote 1).

To access the parameter that maximized the accuracy, you can use this:

```{r}
train_knn$bestTune
```

and the best performing model like this:

```{r}
#| eval: false
train_knn$finalModel
```

The function `predict` will use this best-performing model. Here is the accuracy of the best model when applied to the test set, which we have not yet used because the cross-validation was done on the training set:

```{r}
mean(predict(train_knn, mnist_27$test, type = "raw") == mnist_27$test$y)
```

## Case study: handwritten digit recognition

We are now ready to apply these ideas to a real-world example. Throughout this section, we use the MNIST dataset, introduced in @sec-mnist, as a case study to illustrate the full modeling workflow.

We load the data using the following function from the **dslabs** package:

```{r, message=FALSE, warning=FALSE}
library(dslabs)
mnist <- read_mnist()
```

The dataset includes two components, a training set and a test set:

```{r}
names(mnist)
```

Each of these components includes a matrix with features in the columns:

```{r}
dim(mnist$train$images)
```

and vector with the classes as integers:

```{r}
class(mnist$train$labels)
table(mnist$train$labels)
```

Because we want this example to run on a laptop and in less than one hour, we will consider a subset of the dataset. We will sample 10,000 random rows from the training set and 1,000 random rows from the test set:

```{r}
set.seed(1990)
index <- sample(nrow(mnist$train$images), 10000)
x <- mnist$train$images[index, ]
y <- factor(mnist$train$labels[index])
index <- sample(nrow(mnist$test$images), 1000)
x_test <- mnist$test$images[index, ]
y_test <- factor(mnist$test$labels[index])
```

But you are welcome to try the ideas on the entire dataset.

When fitting models to large datasets, we recommend using matrices instead of data frames, as matrix operations tend to be faster. In the **caret** package, predictor matrices must have column names to track features accurately during prediction on the test set. If the matrices lack column names, you can assign names based on their position:

```{r}
colnames(x) <- 1:ncol(mnist$train$images)
colnames(x_test) <- colnames(x)
```
## Preprocessing

We often transform predictors before running machine learning algorithms. We also remove predictors that are clearly not useful. We call these steps *preprocessing*.

Examples of preprocessing steps include standardizing predictors, applying transformations such as the logarithm, reducing the dimensionality of the feature space, removing highly correlated predictors, and excluding predictors with little variability, for example, those with near-zero variance.

### Feature Filtering

As a first example, we focus on this last type of preprocessing by examining the variability of each predictor. Specifically, we compute the standard deviation of each feature and observe that many have values close to zero, indicating minimal variation:

```{r pixel-sds, message=FALSE, warning=FALSE, cache=FALSE, echo = -1}
rafalib::mypar()
library(matrixStats)
hist(colSds(x), breaks = 256)
```

This is expected because there are parts of the image that rarely contain writing.

The **caret** package includes a function that recommends features to be removed due to *near-zero variance*:

```{r, message=FALSE, warning=FALSE}
nzv <- nearZeroVar(x)
```

We can see the columns recommended for removal are near the edges:

```{r near-zero-image, fig.width = 4, fig.height = 4, out.width="50%", echo=-1}
rafalib::mypar()
image(matrix(1:784 %in% nzv, 28, 28))
```

If we use `nearZeroVar`, we end up removing `length(nzv)` = `r length(nzv)` predictors. We can see that the accuracy of kNN is about the same even after removing all these predictors:

```{r}
mean(predict(knn3(x, y), newdata = x_test, type = "class") == y_test)
mean(predict(knn3(x[, -nzv], y), newdata = x_test[, -nzv], type = "class") == y_test)
```

The **caret** package features the `preProcess` function, which allows users to establish a predefined set of preprocessing operations based on a training set. This function is designed to apply these operations to new datasets without recalculating anything on the validation set, ensuring that all preprocessing steps are consistent and derived solely from the training data.

Below is an example demonstrating how to remove predictors with near-zero variance and then center the remaining predictors:

```{r}
pp <- preProcess(x, method = c("nzv", "center"))
centered_subsetted_x_test <- predict(pp, newdata = x_test)
```

Additionally, the `train` function in caret includes a `preProcess` argument that allows users to specify which preprocessing steps to apply automatically during model training. Here is an example of how to use it to remove near-zero variance predictors. Because this is an illustrative example, we consider only one $k$ and use five-fold cross-validation to estimate the MSE to reduce runtime. As we see in @sec-knn-in-practice, in practice you would want to optimize $k$ and use more thorough resampling:

```{r}
knn_nzv <- train(x, y, method = "knn", preProcess = "nzv", 
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = data.frame(k = 5))
mean(predict(knn_nzv, x_test, type = "raw") == y_test)
```

### Dimension reduction with PCA

An alternative to filtering features directly is to reduce the dimension of the feature matrix using Principal Component Analysis (PCA). For this dataset, a large fraction of the total variability in the predictors can be captured using only a small number of principal components (PCs).

As an example, suppose we want to retain the smallest number of PCs that explain at least 90% of the variability:

```{r}
pca <- prcomp(x)
p <- which(cumsum(pca$sdev^2) / sum(pca$sdev^2) >= 0.9)[1]
```

We can run an algorithm using only these `r p` transformed features. We will use `knn3` from the caret package as an example with the default `k`

```{r}
knn_pca <- knn3(pca$x[, 1:p], y)
```

A critical point when using PCA for prediction is that the PCA transformation must be learned only from the training set. If we use the validation or test set to compute principal components, or even to compute the means used for centering, we inadvertently leak information from those sets into the training process, leading to overfitting.

To avoid this, we compute the necessary centering and rotation matrices on the training set:

```{r}
newdata <- sweep(x_test, 2, colMeans(x)) %*% pca$rotation[, 1:p]
```

The resulting accuracy is similar to what we obtained with the full feature set, but using only `r p` dimensions:

```{r}
mean(predict(knn_pca, newdata, type = "class") == y_test)
```

To obtain an unbiased MSE estimate when training with PCA, we must recompute the PCA inside each cross-validation split and apply the resulting transformation to the corresponding validation fold.

Fortunately, the **caret** package can automate this. The `train` function includes PCA as one of its available preprocessing steps. We can request PCA with a fixed number of components using

`preProcOptions = list(pcaComp = p)`

or ask caret to pick the number of components that explain a fixed fraction of variability, such as 90%, using

`preProcOptions = list(thresh = 0.9)`.

Here is how we modify our earlier code to let caret perform PCA during preprocessing:

```{r}
#| eval: false
knn_pca <- train(x, y, method = "knn", preProcess = c("nzv", "pca"),
  trControl = trainControl("cv", n = 20, preProcOptions = list(thresh = 0.9)))
```

This raises a natural question: why did we choose 0.9? Could another threshold lead to better performance?

One limitation of this approach is that caret does **not** automatically optimize the number of PCs. To tune this quantity directly, just like we tune parameters such as $k$, we would need to define our own model interface. The caret manual^[[https://topepo.github.io/caret/using-your-own-model-in-train.html](https://topepo.github.io/caret/using-your-own-model-in-train.html)] describes how to build custom models that allow full optimization of the PCA dimension.

## Parallelization 

By default, the `train` function estimates model accuracy using 25 bootstrap resamples. This can lead to substantial computation time. For example, if we evaluate 75 different values of a tuning parameter, the model is fit 25 $\times$ 75 = 1,875 times. For algorithms that take several minutes per fit, the total runtime can extend to many hours or even days.

A simple way to approximate the total computation time is to measure how long a single model fit takes using the `system.time` function, and then multiply by the total number of fits. For example, with kNN, we can run:

```{r}
system.time(predict(knn3(x[, -nzv], y), newdata = x_test[, -nzv]))
```

On a typical laptop, this call takes about 2 seconds. This suggests that the full training process could take over an hour.

In practice, this cost can be reduced substantially by using parallel computing.

Imagine you are fitting 100 models; if you had access to 100 computers, you could theoretically speed up the process by a factor of 100 by fitting each model on a separate computer and then aggregating the results. In reality, most modern computers, including many personal computers, are equipped with multiple processors that allow for such parallel execution. This method, known as parallelization, leverages these multiple processors to conduct several computational tasks simultaneously, significantly accelerating the model training process. By distributing the workload across different processors, parallelization makes it feasible to manage large datasets and complex modeling procedures efficiently.

The **caret** package is designed to take advantage of parallel processing, but you need to explicitly tell R to run tasks in parallel. To do this, we use the **doParallel** package. We will see how to set up parallelization for the kNN example in the next section, but the same approach can be applied to any algorithm that uses the `train` function.

:::{.callout-warning}
When parallelizing tasks across multiple processors, it is important to consider the risk of running out of memory. Each processor might require a copy of the data or substantial portions of it, which can multiply overall memory demands. This is especially challenging if the data or models are large.
:::

## Model tuning and diagnostics {#sec-knn-in-practice}

In Sections [-@sec-knn-cv-intro] and [-@sec-random-forests], we introduced the k-nearest neighbors (kNN) and random forest algorithms, along with their intuition and formulation. In this section, we use these two methods as working examples to illustrate the process of model selection and tuning. Specifically, we demonstrate how to train and evaluate models in R, choose appropriate hyperparameters, such as $k$ for kNN and tree-related parameters for random forests, and assess performance using cross-validation.


### Optimizing tuning parameters

We start by demonstrating how we select the tuning parameter $k$ in the kNN algorithm. In **caret**, this is done using the `tuneGrid` argument in the `train` function. The grid of candidate values must be provided as a data frame, with column names matching those returned by `modelLookup`. For kNN, this means a column named `k`. The default values are 5, 7, and 9, which is a very limited range.

In practice, selecting the grid and resampling strategy involves a tradeoff. Using fewer cross-validation folds reduces computation time but yields noisier estimates of prediction error, while more folds provide more precise estimates at greater runtime cost.

Because of this tradeoff, it is inefficient to search over a fine grid from the start. Instead, we begin with a coarse grid over a broad range and a faster resampling scheme to identify a promising region. Once we have a rough idea of where the optimal $k$ lies, we refine the search with a finer grid and more thorough resampling.

Throughout this process, we use parallel computation to reduce runtime, since each combination of tuning parameters and resamples can be evaluated independently.

```{r mnist-knn-fit}
#| message: false
#| warning: false
library(doParallel)
nc <- detectCores() - 1   # it is convention to leave 1 core for the OS
cl <- makeCluster(nc)
registerDoParallel(cl)
set.seed(1991)
knn <- train(x, y, method = "knn", preProcess = c("nzv", "pca"),
                   trControl = trainControl(method = "cv", number = 5,
                           preProcOptions = list(thresh = 0.9)),
                   tuneGrid = data.frame(k = seq(1, 25, 4)))
plot(knn, highlight = TRUE)
```

Based on these results, we can infer that the best $k$ is between 1 and 5. We now refine the search with a finer grid and 20-fold cross-validation:

```{r}
set.seed(1992)
knn <- train(x, y, method = "knn", preProcess = c("nzv", "pca"),
                   trControl =  trainControl(method = "cv", number = 20,
                           preProcOptions = list(thresh = 0.9)),
                   tuneGrid = data.frame(k = seq(1, 7, 2)))
knn$bestTune
```

to choose $k$ = `r knn$bestTune$k`. Now that we've optimize our algorithm, the `predict` function defaults to using the best-performing model fit with the entire training data:

```{r}
y_hat_knn <- predict(knn, x_test, type = "raw")
```

We achieve relatively high accuracy:

```{r}
mean(y_hat_knn == y_test)
```

Once done with parallelization, it is good practice to stop the cluster:

```{r eval = FALSE}
stopCluster(cl)
stopImplicitCluster()
```


## Improving computational efficiency

With the random forest algorithm, several parameters can be optimized, but the main one is `mtry`, the number of predictors that are randomly selected for each tree. This is also the only tuning parameter that the **caret** function `train` permits when using the default implementation from the **randomForest** package.

```{r message = FALSE}
#| cache: false
library(randomForest)
```

The random forest algorithm is computationally more intensive than kNN and can require substantially longer training times. As a result, tuning its hyperparameters requires a more deliberate strategy. In this example, we adopt several approaches to reduce computation time: we use parallelization, decrease the number of bootstrap resamples, limit the number of trees grown in each fit, and perform an initial search over a coarse grid before refining the search in a narrower region.

We start by fitting the default model and checking 


```{r accuracy-versus-ntree, echo=-1}
rafalib::mypar()
rf <- randomForest(x[, -nzv], y)
plot(rf)
```
We see that accuracy begins to level off at around 200 trees. So for our coarse search, we limit the number of trees to 200, use five-fold cross-validation instead of 25 bootstrap samples, and consider a coarse grid of `mtry` values. This reduction in the number of trees and the size of the resampling allows us to reduce the runtime by a factor of 10, which allows us to identify a promising region for `mtry` relatively quickly:

```{r rf-coarse}
#| warning: false
library(doParallel)
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
set.seed(1993)
rf <- train(x, y, method = "rf", ntree = 250, preProcess = "nzv",
  trControl = trainControl(method = "cv", n = 5), 
  tuneGrid = data.frame(mtry = seq(3, 53, 10)))
ggplot(rf, highlight = TRUE)
```

We see that the optimal `mtry` is between 5 and 20. 

::: {.callout-note}
Note that with smaller values of `mtry`, individual trees tend to be weaker because each split considers fewer predictors. As a result, more trees are typically required for performance to stabilize. In contrast, larger values of `mtry` produce stronger but more correlated trees, so fewer trees may suffice. 

When reducing the number of trees to speed up tuning, keep this tradeoff in mind: using too few trees can bias comparisons across `mtry` values, particularly disadvantaging smaller `mtry`. It is therefore advisable to use enough trees to obtain stable performance when selecting the range of `mtry`.
:::

We now refine the search using a finer grid and the default resampling settings. Note that the code below uses 500 trees and 25 bootstrap resamples, so it may take several minutes to run.

```{r, eval=FALSE}
set.seed(1994)
rf <- train(x, y, method = "rf", preProcess = "nzv", 
  tuneGrid = data.frame(mtry = 5:20))
```

Now that we have optimized our algorithm, we are ready to predict outcomes with our final model:

```{r}
#| echo: false
#| message: false
## we hard wire to make book compilation faster.
## we ran previous code once to determine 9 was the best
library(randomForest)
nzv <- nearZeroVar(x)
fit_rf <- randomForest(x[, -nzv], y, mtry = 6)
y_hat_rf <- predict(fit_rf, x_test[, -nzv])
```


```{r}
#| eval: false
y_hat_rf <- predict(train_rf, x_test, type = "raw")
```


As with kNN, we also achieve high accuracy:

```{r}
mean(y_hat_rf == y_test)
```

By optimizing some of the other algorithm parameters, we can achieve even higher accuracy.

To end, we turn off the use of parallel:

```{r}
stopCluster(cl)
stopImplicitCluster()
```

:::{.callout-note}
## Variable importance

We described variable importance in @sec-variable-importance. The following function computes the importance of each feature:

```{r}
#| eval: false
imp <- varImp(rf)$importance[,1]
```
```{r}
#| echo: false
imp <- varImp(fit_rf)$Overall
```

We can see which features are being used most by plotting an image:

```{r importance-image, echo=-1, fig.width = 4, fig.height = 4, out.width="50%"}
rafalib::mypar()
mat <- rep(0, ncol(x))
mat[-nzv] <- imp
image(matrix(mat, 28, 28))
```
:::

## Diagnostics

An important part of data analysis is visualizing results to determine why we are failing. How we do this depends on the application. Below we show the images of digits for which we made an incorrect prediction. Here are some errors for the random forest:

```{r rf-images, echo=FALSE, out.width="100%", fig.width=6, fig.height=1.65}
p_max <- predict(rf, x_test[, -nzv], type = "prob")
p_max <- p_max / rowSums(p_max)
p_max <- apply(p_max, 1, max)

ind <- which(y_hat_rf != y_test)
ind <- ind[order(p_max[ind], decreasing = TRUE)]

rafalib::mypar(1, 4)
for (i in ind[1:4]) {
  image(
    matrix(x_test[i, ], 28, 28)[, 28:1], 
      main = paste0("Pr(", y_hat_rf[i], ")=", round(p_max[i], 2), " but is a ", y_test[i]),
    xaxt = "n", yaxt = "n")
}
```

By examining errors like this, we often find specific weaknesses in algorithms or parameter choices and can try to correct them.

## Ensembles

The idea of an ensemble is similar to the idea of combining data from different pollsters to obtain a better estimate of the true support for each candidate.

In machine learning, one can usually greatly improve the final results by combining the results of different algorithms.

Here is a simple example where we compute new class probabilities by taking the average of the previously fitted random forest (saved in `rf`) and kNN (saved in `knn`). We can see that the accuracy improves:

```{r}
p_rf <- predict(rf, x_test[, -nzv], type = "prob")
p_rf <- p_rf / rowSums(p_rf)
p_knn <- predict(knn, x_test, type = "prob")
p <- (p_rf + p_knn) / 2
y_pred <- factor(apply(p, 1, which.max) - 1)
mean(y_pred == y_test)
```

We have just built an ensemble with just two algorithms. By combining more similarly performing but uncorrelated algorithms, we can improve accuracy further.

## Exercises


1\. Create a simple dataset where the outcome grows 0.75 units on average for every increase in a predictor: 

```{r forest-q1}
n <- 1000
sigma <- 0.25
x <- rnorm(n, 0, 1)
y <- 0.75 * x + rnorm(n, 0, sigma)
dat <- data.frame(x = x, y = y)
```

In the exercises in @sec-example-algorithms, we saw that changing `maxnodes` or `nodesize` in the `randomForest` function improved our estimate in this dataset. Let's use the `train` function to help us pick these values. From the **caret** manual, we see that we can't tune the `maxnodes` parameter or the `nodesize` argument with `randomForest`, so we will use the **Rborist** package and tune the `minNode` argument. Use the `train` function to try values `minNode <- seq(5, 250, 25)`. See which value minimizes the estimated RMSE.

2\. Make a scatterplot of `x` and `y` and add  the prediction from the best-fitted model in exercise 1.

3\. This **dslabs** `tissue_gene_expression` dataset includes a matrix `x`:

```{r, eval = FALSE}
library(dslabs)
dim(tissue_gene_expression$x)
```

with the gene expression measured on 500 genes for 189 biological samples representing seven different tissues. The tissue type is stored in `y`:

```{r, eval = FALSE}
table(tissue_gene_expression$y)
```

Split the data into training and test sets, then use kNN to predict tissue type and see what accuracy you obtain. Try it for $k = 1, 3, \dots, 11$.

4\. We are going to apply LDA and QDA to the `tissue_gene_expression` dataset. We will start with simple examples based on this dataset and then develop a realistic example.

Create a dataset with just the classes `cerebellum` and `hippocampus` (two parts of the brain) and a predictor matrix with 10 randomly selected columns. 


```{r, eval = FALSE}
set.seed(1993)
tissues <- c("cerebellum", "hippocampus")
ind <- which(tissue_gene_expression$y %in% tissues)
y <- droplevels(tissue_gene_expression$y[ind])
x <- tissue_gene_expression$x[ind, ]
x <- x[, sample(ncol(x), 10)]
```

Estimate the accuracy of LDA.

5\. In this case, LDA fits two 10-dimensional normal distributions. Look at the fitted model by looking at the `finalModel` component of the result of train. Notice there is a component called `means` that includes the estimated `means` of both distributions. Plot the mean vectors against each other and determine which predictors (genes) appear to be driving the algorithm. 

6\. Repeat exercises 4 with QDA. Does it have a higher accuracy than LDA?

7\. Are the same predictors (genes) driving the algorithm? Make a plot as in exercise 5.

8\. One thing we see in the previous plot is that predictors are correlated between groups: some predictors are low in both groups, while others are high in both groups. The mean value of each predictor, `colMeans(x)`, is not informative or useful for prediction, and often it is useful to center or scale each column for interpretation purposes. This can be achieved with the `preProcessing` argument in `train`. Re-run LDA with `preProcessing = "scale"`. Note that accuracy does not change, but see how it is easier to identify the predictors that differ more between groups than the plot made in exercise 5.

9\. In the previous exercises, we saw that both approaches worked well. Plot the predictor values for the two genes with the largest differences between the two groups in a scatterplot to see how they appear to follow a bivariate distribution as assumed by the LDA and QDA approaches. Color the points by the outcome.

10\. Now we are going to increase the complexity of the challenge slightly: we will consider all the tissue types.

```{r, eval = FALSE}
set.seed(1993)
y <- tissue_gene_expression$y
x <- tissue_gene_expression$x
x <- x[, sample(ncol(x), 10)]
```

What accuracy do you get with LDA?


11\. We see that the results are slightly worse. Use the `confusionMatrix` function to learn what type of errors we are making.


12\. Plot an image of the centers of the seven 10-dimensional normal distributions used by the LDA model fit in exercise 10.


13\. Use the `rpart` function to fit a classification tree to the `tissue_gene_expression` dataset with all genes. That is, redefine 

```{r}
#| eval: false
x <- tissue_gene_expression$x
```

Use the `train` function to estimate the accuracy. Try out `cp` values of `seq(0, 0.05, 0.01)`. Plot the accuracy to report the results of the best model.


14\. Study the confusion matrix for the best-fitting classification tree. What do you observe happening for placenta samples?

15\. Notice that placentas are called endometrium more often than placenta. Note also that there are only six placenta samples, and that, by default, `rpart` requires 20 observations before splitting a node. Thus, it is not possible with these parameters to have a node in which placentas are the majority. Rerun the above analysis, but this time permit `rpart` to split any node by using the argument `control = rpart.control(minsplit = 0)`. Does the accuracy increase? Look at the confusion matrix again.

16\. Plot the tree from the best-fitting model obtained in exercise 13.

17\. We can see that with just the genes selected by CART, we are able to predict the tissue type. Now let's see if we can do even better with a random forest. Use the `train` function in the **caret** package and the `rf` method to train a random forest. Try out values of `mtry` ranging from, at least, `seq(50, 200, 25)`. What `mtry` value maximizes accuracy? To permit small `nodesize` to grow as we did with the classification trees, use the following argument: `nodesize = 1`.  This will take several seconds to run. If you want to test it out, try using smaller values with `ntree`. Set the seed to 1990.

18\. Use the function `varImp` on the output of `train` and save it to an object called `imp`.

19\. The `rpart` model we ran in exercise 15 produced a tree that used just a few predictors. Extracting the predictor names is not straightforward, but can be done. If the output of the call to train was `fit_rpart`, we can extract the names like this:

```{r, eval = FALSE}
ind <- !(fit_rpart$finalModel$frame$var == "<leaf>")
tree_terms <-
  fit_rpart$finalModel$frame$var[ind] |>
  unique() |>
  as.character()
tree_terms
```

What is the variable importance in the random forest call for these predictors? Where do they rank?

20\. Extract the top 50 predictors based on importance, take a subset of `x` with just these predictors, and apply the function `heatmap` to see how these genes behave across the tissues. We will introduce the `heatmap` function in @sec-clustering.


21\. In @sec-example-algorithms, for the 2 or 7 example, we compared the conditional probability $p(\mathbf{x})$ given two predictors $\mathbf{x} = (x_1, x_2)^\top$ to the fit $\hat{p}(\mathbf{x})$ obtained with a machine learning algorithm by making image plots. The following code can be used to make these images and include a curve at the values of $x_1$ and $x_2$ for which the predicted conditional probability is $0.5$:

```{r}
#| eval: false
plot_cond_prob <- function(x_1, x_2, p) {
  data.frame(x_1 = x_1, x_2 = x_2, p = p) |>
    ggplot(aes(x_1, x_2)) +
    geom_raster(aes(fill = p), show.legend = FALSE) +
    stat_contour(aes(z = p), breaks = 0.5, color = "black") +
    scale_fill_gradientn(colors = c("#F8766D", "white", "#00BFC4"))
}
```

We can see the true conditional probability for the 2 or 7 example like this:

```{r}
#| eval: false
with(mnist_27$true_p, plot_cond_prob(x_1, x_2, p))
```


Fit a kNN model and make this plot for the estimated conditional probability. Hint: Use the argument `newdata = mnist_27$train` to obtain predictions for a grid of points.

22\. Notice that, in the plot made in exercise 21, the boundary is somewhat wiggly. This is because kNN, like the basic bin smoother, does not use a kernel. To improve this, we could try loess. By reading through the available models part of the **caret** manual, we see that we can use the `gamLoess` method. We need to install the __gam__ package, if we have not done so already. We see that we have two parameters to optimize:

```{r}
modelLookup("gamLoess")
```

Use resampling to pick a span between 0.15 and 0.75. Keep `degree = 1`. What span is selected?


23\. Show an image plot of the estimate $\hat{p}(x,y)$ resulting from the model fit in exercise 22. How does the accuracy compare to that of kNN? Comment on the difference between the estimate obtained with kNN. 

24\. Use the `mnist_27` training set to fit a variety of models available through the **caret** package.
Your goal is simply to *try many different algorithms* and compare their performance, even if we have not explained them yet.

Here is a vector of model names you can try:

```{r}
models <- c(
  "glm",
  "lda",
  "naive_bayes",
  "svmLinear",
  "gamboost",
  "gamLoess",
  "qda",
  "knn",
  "kknn",
  "loclda",
  "gam",
  "rf",
  "ranger",
  "wsrf",
  "Rborist",
  "avNNet",
  "mlp",
  "monmlp",
  "gbm",
  "adaboost",
  "svmRadial",
  "svmRadialCost",
  "svmRadialSigma"
)
```

Fit each model using `train()` with default tuning parameters and store the fitted model objects in a list.
You may need to install additional packages depending on which models you try, and some models will produce warnings—that’s expected.

25\. Now that you have all the trained models in a list, use `sapply` or `map` to create a matrix of predictions for the test set. You should end up with a matrix with `length(mnist_27$test$y)` rows and `length(models)` columns.

26\. Compute accuracy for each model on the test set.

27\. Build an ensemble prediction by majority vote and compute the accuracy of the ensemble.

28\. Earlier, we computed the accuracy of each method on the training set and noticed that they varied. Which individual methods do better than the ensemble?

29\. It is tempting to remove the methods that do not perform well and redo the ensemble. The problem with this approach is that we are using the test data to make a decision. However, we could use the accuracy estimates obtained from cross-validation with the training data. Obtain these estimates and save them in an object.

30\. Now, let's only consider the methods with an estimated accuracy of 0.8 when constructing the ensemble. What is the accuracy of the ensemble now?

31\. Note that if two machine learning algorithms predict the same exact outcome, ensembling them will not change the prediction. For each pair of algorithms, compare the percent of observations for which they make the same prediction. Use the `heatmap` function to visualize the results. Hint: use the `method = "binary"` argument in the `dist` function.

32\. Note that each method can also produce an estimated conditional probability. For most methods, we can use `type = "prob"` in the predict function. Note that some of the methods require you to use the argument `trControl = trainControl(classProbs=TRUE)` when calling train. Also, these methods do not work if classes have numbers as names. Hint: change the levels like this:

```{r, eval = FALSE}
dat$train$y <- recode_factor(dat$train$y, "2" = "two", "7" = "seven")
dat$test$y <- recode_factor(dat$test$y, "2" = "two", "7" = "seven")
```

Build an ensemble predicted conditional probabilities by taking the average of these estimated conditional probabilities instead of a majority vote. Determine class predictions with a conditional probability cutoff of 0.5. What is the accuracy of this new ensemble?

33\. In this chapter, we illustrated a couple of machine learning algorithms on a subset of the MNIST dataset. Try fitting a model to the entire dataset.

34\. We fit the following kNN model to the MNIST data:

```{r}
#| eval: false
knn_nzv <- train(x, y, method = "knn", preProcess = "nzv", 
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = data.frame(k = 5))
```

Fit it again, but this time use 20-fold cross-validation instead of 5-fold and search over a wider range of $k$ values, for example, `tuneGrid = data.frame(k = seq(3, 7, 2))`. Do this with and without `preProcess = "nzv"` and compare the results.
