# Machine learning in practice {#sec-ml-in-practice}

Now that we have learned several methods and explored them with simple examples, we will try them out on a real example: the MNIST digits.

We can load this data using the following **dslabs** package:

```{r}
#| cache: false
#| echo: false
library(dslabs)
```

```{r, message=FALSE, warning=FALSE}
library(dslabs)
mnist <- read_mnist()
```

The dataset includes two components, a training set and a test set:

```{r}
names(mnist)
```

Each of these components includes a matrix with features in the columns:

```{r}
dim(mnist$train$images)
```

and vector with the classes as integers:

```{r}
class(mnist$train$labels)
table(mnist$train$labels)
```

Because we want this example to run on a small laptop and in less than one hour, we will consider a subset of the dataset. We will sample 10,000 random rows from the training set and 1,000 random rows from the test set:

```{r}
set.seed(1990)
index <- sample(nrow(mnist$train$images), 10000)
x <- mnist$train$images[index,]
y <- factor(mnist$train$labels[index])
index <- sample(nrow(mnist$test$images), 1000)
x_test <- mnist$test$images[index,]
y_test <- factor(mnist$test$labels[index])
```

## The caret package {#sec-caret}

We have already learned about several machine learning algorithms. Many of these algorithms are implemented in R. However, they are distributed via different packages, developed by different authors, and often use different syntax. The __caret__ package tries to consolidate these differences and provide consistency. It currently includes over 200 different methods which are summarized in the __caret__ package manual^[https://topepo.github.io/caret/available-models.html]. Keep in mind that __caret__ does not include the packages needed to run each possible algorithm. To apply a machine learning method through __caret__ you still need to install the library that implement the method. The required packages for each method are described in the package manual. 

The __caret__ package also provides a function that performs cross validation for us. Here we provide some examples showing how we use this helpful package. We will first use the 2 or 7 example to illustrate and, in later sections, we use the package to run algorithms on the larger MNIST dataset.

### The `train` function 

The R functions that fit machine algorithms are all slightly different. Functions such as `lm`, `glm`, `qda`, `lda`, `knn3`, `rpart` and `randomForrest` use different syntax, have different argument names and produce objects of different types.

The __caret__ `train` function lets us train different algorithms using similar syntax. So, for example, we can type:

```{r}
#| cache: false
library(caret)
train_glm <- train(y ~ ., method = "glm", data = mnist_27$train)
train_qda <- train(y ~ ., method = "qda", data = mnist_27$train)
train_knn <- train(y ~ ., method = "knn", data = mnist_27$train)
```

FIX As we explain in more detail in #sec-caret-cv, the train function selects parameters for you using cross validation.

### The `predict` function

The `predict` function is very useful for machine learning applications. This function takes an object from a fitting function and a data frame with features \mathbf{x} for which to predict, and returns predictions for these features. 

Here is an example with logistic regression:

```{r}
fit <- glm(y ~ ., data = mnist_27$train, family = "binomial")
p_hat <- predict(fit, mnist_27$test)
```

In this case, the function is simply computing: 

$$
\hat{p}(\mathbf{x}) = g^{-1}\left(\hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2 \right) \text{ with } g(p) = \log\frac{p}{1-p} \implies g^{-1}(\mu) = \frac{1}{1-e^{-\mu}}
$$

for the `x_1` and `x_2` in the test set `mnist_27$test`. With these estimates in place, we can make our predictions and compute our accuracy:

```{r}
y_hat <- factor(ifelse(p_hat > 0.5, 7, 2))
```

However, note that `predict` does not always return objects of the same types; it depends on what type of object it is applied to. To learn about the specifics, you need to look at the help file specific for the type of fit object that is being used. The `predict` is actually a special type of function in R (called a _generic function_) that calls other functions depending on what kind of object it receives. So if `predict` receives an object coming out of the `lm` function, it will call `predict.glm`. If it receives an object coming out of `glm`, it calls `predict.qda`. If the fit is from `knn3`, it calls `predict.knn3`, and so on. These functions are similar but not exactly. You can learn more about the differences by reading the help files:

```{r, eval = FALSE}
?predict.glm
?predict.qda
?predict.knn3
```

There are many other versions of `predict` and many machine learning algorithms have a `predict` function.

As with `train`, the **caret** packages unifies the use of `predict` with the function `predict.train`. This function takes the output of `train` and produces prediction of categories or estimates of $p(\mathbf{x}).

The code looks the same for all methods:

```{r}
y_hat_glm <- predict(train_glm, mnist_27$test, type = "raw")
y_hat_qda <- predict(train_qda, mnist_27$test, type = "raw")
y_hat_knn <- predict(train_knn, mnist_27$test, type = "raw")
```

This permits us to quickly compare the algorithms. For example, we can compare the accuracy like this:

```{r}
fits <- list(glm = y_hat_glm, qda = y_hat_qda, knn = y_hat_knn)
sapply(fits, function(fit) confusionMatrix(fit, mnist_27$test$y)$overall[["Accuracy"]])
```


### Cross validation {#sec-caret-cv}

When an algorithm includes a tuning parameter, `train` automatically uses cross validation to decide among a few default values. To find out what parameter or parameters are optimized, you can read the manual ^[http://topepo.github.io/caret/available-models.html] or study the output of: 

```{r, eval=FALSE}
getModelInfo("knn")
```

We can also use a quick lookup like this:

```{r, eval=FALSE}
modelLookup("knn")
```

If we run it with default values: 

```{r}
train_knn <- train(y ~ ., method = "knn", data = mnist_27$train)
```

you can quickly see the results of the cross validation using the `ggplot` function. The argument `highlight` highlights the max:

```{r caret-highlight}
ggplot(train_knn, highlight = TRUE)
```

By default, the cross validation is performed by taking 25 bootstrap samples comprised of 25% of the observations. For the `kNN` method, the default is to try $k=5,7,9$. We change this using the `tuneGrid` argument. The grid of values must be supplied by a data frame with the parameter names as specified in the `modelLookup` output. 

Here we present an example where we try out 30 values between 9 and 67. To do this with __caret__, we need to define a column named `k`, so we use this:  `data.frame(k = seq(9, 67, 2))`. Note that when running this code, we are fitting 30 versions of kNN to 25 bootstrapped samples. Since we are fitting $30 \times 25 = 750$ kNN models, running this code will take several seconds. We set the seed because cross validation is a random procedure and we want to make sure the result here is reproducible.

```{r train-knn-plot}
set.seed(2008)
train_knn <- train(y ~ ., method = "knn", 
                   data = mnist_27$train,
                   tuneGrid = data.frame(k = seq(9, 71, 2)))
ggplot(train_knn, highlight = TRUE)
```

To access the parameter that maximized the accuracy, you can use this:

```{r}
train_knn$bestTune
```

and the best performing model like this:

```{r}
train_knn$finalModel
```

The function `predict` will use this best performing model. Here is the accuracy of the best model when applied to the test set, which we have not yet used because the cross validation was done on the training set:

```{r}
confusionMatrix(predict(train_knn, mnist_27$test, type = "raw"),
                mnist_27$test$y)$overall["Accuracy"]
```

If we want to change how we perform cross validation, we can use the `trainControl` function. We can make the code above go a bit faster by using, for example, 10-fold cross validation. This means we have 10 samples using 10% of the observations each. We accomplish this using the following code:

```{r cv-10-fold-accuracy-estimate}
control <- trainControl(method = "cv", number = 10, p = .9)
train_knn_cv <- train(y ~ ., method = "knn", 
                   data = mnist_27$train,
                   tuneGrid = data.frame(k = seq(9, 71, 2)),
                   trControl = control)
ggplot(train_knn_cv, highlight = TRUE)
```

We observe that the accuracy estimates are more variable, which is expected since we changed the number of samples used to estimate accuracy.

Note that `results` component of the `train` output includes several summary statistics related to the variability of the cross validation estimates:

```{r}
names(train_knn$results)
```

You can learn many more details about the **caret** package, from the manual ^[https://topepo.github.io/caret/available-models.html].

## Preprocessing

We often transform predictors before running the machine algorithm. We also remove predictors that are clearly not useful. We call these steps *preprocessing*.

Examples of preprocessing include standardizing the predictors, taking the log transform of some predictors, removing predictors that are highly correlated with others, and removing predictors with very few non-unique values or close to zero variation. FIX DO YOU NEED THIS SENT: We show an example below.

For example, we can run the `nearZero` function from the **caret** package to see that several features do not vary much from observation to observation. We can see that there is a large number of features with 0 variability:

```{r pixel-sds, message=FALSE, warning=FALSE, cache=FALSE}
library(matrixStats)
sds <- colSds(x)
hist(sds, breaks = 256)
```

This is expected because there are parts of the image that rarely contain writing (dark pixels).

The **caret** packages includes a function that recommends features to be removed due to *near zero variance*:

```{r, message=FALSE, warning=FALSE}
nzv <- nearZeroVar(x)
```

We can see the columns recommended for removal:

```{r, eval=FALSE}
image(matrix(1:784 %in% nzv, 28, 28))
```

```{r near-zero-image, fig.width = 4, fig.height = 4, out.width="50%"}
rafalib::mypar()
image(matrix(1:784 %in% nzv, 28, 28))
```

So we end up keeping this number of columns:

```{r}
col_index <- setdiff(1:ncol(x), nzv)
length(col_index)
```

Now we are ready to fit some models. Before we start, we need to add column names to the feature matrices as these are required by **caret**:

```{r}
colnames(x) <- 1:ncol(mnist$train$images)
colnames(x_test) <- colnames(x)
```

## k-nearest neighbors

:::{.callout-warning}
Before starting this section, note that the first two calls to the `train` function in the code below can take several hours to run. This is a common challenge when training machine learning algorithms since we have to run the algorithm for each cross validation split and each set of tuning parameters being considered. In the next section, we will provide some suggestions on how to predict the duration of the process and ways to reduce. 
:::

The first step is to optimize for $k$.

```{r mnist-knn-fit, eval=FALSE}
train_knn <- train(x[ ,col_index], y, 
                   method = "knn", 
                   tuneGrid = data.frame(k = seq(3, 13, 2)))
```

Once we optimize our algorithm, we can fit it to the entire dataset:

```{r, eval = FALSE}
fit_knn <- knn3(x[, col_index], y,  k = train_knn$bestTune)
```

```{r, echo = FALSE}
fit_knn <- knn3(x[, col_index], y,  k = 7)
```

We achieve a high accuracy:

```{r}
y_hat_knn <- predict(fit_knn, x_test[, col_index], type = "class")
confusionMatrix(y_hat_knn, factor(y_test))$overall["Accuracy"]
```

### Dimension reduction with PCA

An alternative to removing low variance columns directly is to use dimension reduction on the feature matrix before applying the algorithms. It is important that we not use the test set when finding the PCs nor any summary of the data, as this could result in overtraining. So we start by applying `prcomp` to the training data:

```{r}
col_means <- colMeans(x)
pca <- prcomp(sweep(x, 2, col_means), center = FALSE)
```

Next, we run kNN on just a small number of dimensions. We try 36 dimensions since this explains about 80% of the data. 

```{r}
#| eval: false

k <- 36
x_train <- pca$x[,1:k]
train_knn <- train(x_train, y, 
                   method = "knn", 
                   tuneGrid = data.frame(k = seq(3, 13, 2)))
fit_knn_pca <- knn3(x_train, y, k = train_knn$bestTune)
```

```{r}
#| echo: false

k <- 36
fit_knn_pca <- knn3(pca$x[,1:k], y, k = 7)
```

Now we apply the transformation we learned with the training data to the test data, reduce the dimension, and then run predict. Note that we used the rotation and column means estimated from the training data.

```{r}
newdata <-  sweep(x_test, 2, col_means) %*% pca$rotation[,1:k]
y_hat_knn_pca <- predict(fit_knn_pca, newdata, type = "class")
confusionMatrix(y_hat_knn_pca, factor(y_test))$overall["Accuracy"]
```

FIX With obtain an improvement in accuracy, while using only 36 dimensions. 

:::{.callout-note}
Remember the entire algorithm needs to be developed on the training data. FIX This is why the column means subtracted from test set columns were those estimated in the training set and the rotation applied to the test data to obtained PCs we the one obtained with the train data. 
:::


## Random Forest

With the random forest algorithm several parameters can be optimized, but the main one is `mtry`, the number of predictors that are randomly selected for each tree. This is also the only tuning parameter that the **caret** function `train` permits when using the default implementation from the **randomForest** package.

```{r, eval=FALSE}
library(randomForest)
grid <- data.frame(mtry = seq(3, 24, 3))
train_rf <-  train(x[, col_index], y, 
                   method = "rf", 
                   tuneGrid = grid)
```

Now that we have optimized our algorithm, we are ready to fit our final model:

```{r, eval=FALSE}
fit_rf <- randomForest(x[, col_index], y, mtry = train_rf$bestTune$mtry)
```

```{r}
#| echo: false
#| cache: false
#| message: false
library(randomForest)
```

```{r, echo=FALSE}
## we hard wire to make book compilation faster.
## we ran previous code once to determine 9 was the best
fit_rf <- randomForest(x[, col_index], y,  mtry = 9)
```

As with kNN, we also achieve high accuracy:

```{r}
y_hat_rf <- predict(fit_rf, x_test[ ,col_index])
confusionMatrix(y_hat_rf, y_test)$overall["Accuracy"]
```

By optimizing some of the other algorithm parameters, we can achieve even higher accuracy.

## Testing and improving computation time

The default method for estimating accuracy used by the `train` function is to test prediction on 25 bootstrap samples. This can result in long compute times. For example, if we are considering several values, say 10, of the tuning parameters, we will fit the algorithm 250 times. We can use the `system.time` function to estimate how long it takes to run the algorithm once:

```{r}
system.time({fit_rf <- randomForest(x[, col_index], y,  mtry = 9)})
```

and use this to estimate the total time for the 250 iterations. In this case it will be several hours.

One way to reduce run time is to use k-fold cross validation with a smaller number of test sets. A popular choice is leaving out 10 test sets with 10% of the data:

```{r mnist-knn-fit-control, eval=FALSE}
control <- trainControl(method = "cv", number = 10, p = .9)
```

and re-running the `train` function with this choice specified via the `trControl` argument:

```{r, eval=FALSE}
train_rf <-  train(x[, col_index], y, 
                   method = "rf", 
                   tuneGrid = grid,
                   trControl = control)
```

For random forest, we can also speed up the training step by running less trees per fit. After running the algorithm once, we can use the plot function to see how the error rate changes as the number of trees grows. 

Here we can see that error rate stabilizes after about 200 trees:

```{r}
plot(fit_rf)
```

We can use this finding to speed up the cross validation procedure. Specifically, because the default is 500, by adding the argument `ntree = 200` to the call to `train` above, the procedure will finish 2.5 times faster.

## Variable importance

The following function computes the importance of each feature:

```{r}
imp <- importance(fit_rf)
```

We can see which features are being used most by plotting an image:

```{r eval=FALSE}
mat <- rep(0, ncol(x))
mat[col_index] <- imp
image(matrix(mat, 28, 28))
```

```{r importance-image, echo=FALSE, fig.width = 4, fig.height = 4, out.width="50%"}
rafalib::mypar()
mat <- rep(0, ncol(x))
mat[col_index] <- imp
image(matrix(mat, 28, 28))
```

## Diagnostics

An important part of data analysis is visualizing results to determine why we are failing. How we do this depends on the application. Below we show the images of digits for which we made an incorrect prediction. Here are some errors for the random forest:

```{r rf-images,, echo=FALSE, out.width="100%", fig.width=6, fig.height=1.65}
p_max <- predict(fit_rf, x_test[,col_index], type = "prob") 
p_max <- p_max / rowSums(p_max)
p_max <- apply(p_max, 1, max)

ind  <- which(y_hat_rf != y_test)
ind <- ind[order(p_max[ind], decreasing = TRUE)]

rafalib::mypar(1,4)
for (i in ind[1:4]) {
  image(matrix(x_test[i,], 28, 28)[, 28:1], 
        main = paste0("Pr(",y_hat_rf[i],")=",round(p_max[i], 2), " but is a ",y_test[i]),
        xaxt = "n", yaxt = "n")
}
```

By examining errors like this, we often find specific weaknesses to algorithms or parameter choices and can try to correct them.

## Ensembles

The idea of an ensemble is similar to the idea of combining data from different pollsters to obtain a better estimate of the true support for each candidate.

In machine learning, one can usually greatly improve the final results by combining the results of different algorithms.

Here is a simple example where we compute new class probabilities by taking the average of random forest and kNN. We can see that the accuracy improves:

```{r}
p_rf <- predict(fit_rf, x_test[,col_index], type = "prob")  
p_rf <- p_rf / rowSums(p_rf)
p_knn_pca  <- predict(fit_knn_pca, newdata)
p <- (p_rf + p_knn_pca)/2
y_pred <- factor(apply(p, 1, which.max) - 1)
confusionMatrix(y_pred, y_test)$overall["Accuracy"]
```

We have just built an ensemble with just two algorithms. By combing more similarly performing, but uncorrelated, algorithms we can improve accuracy further.

## Exercises

1\. In the exercises in @sec-example-alogirhms we saw that changing `maxnodes` or `nodesize` in the `randomForest` function improved our estimate. Let's use the `train` function to help us pick these values. From the **caret** manual we see that we can't tune the `maxnodes` parameter or the `nodesize` argument with `randomForest`, so we will use the **Rborist** package and tune the `minNode` argument. Use the `train` function to try values `minNode <- seq(5, 250, 25)`. See which value minimizes the estimated RMSE.

2\. This **dslabs* dataset includes a matrix `x`:

```{r, eval = FALSE}
library(dslabs)
dim(tissue_gene_expression$x)
```

with the gene expression measured on 500 genes for 189 biological samples representing seven different tissues. The tissue type is stored in `y`:

```{r, eval = FALSE}
table(tissue_gene_expression$y)
```

Split the data in training and test sets, then use kNN to predict tissue type and see what accuracy you obtain. Try it for $k = 1, 3, \dots, 11$.

3\. We are going to apply LDA and QDA to the `tissue_gene_expression` dataset. We will start with simple examples based on this dataset and then develop a realistic example.

Create a dataset with just the classes `cerebellum` and `hippocampus` (two parts of the brain) and a predictor matrix with 10 randomly selected columns. Estimate the accuracy of LDA.


```{r, eval = FALSE}
set.seed(1993)
tissues <- c("cerebellum", "hippocampus")
ind <- which(tissue_gene_expression$y %in% tissues)
y <- droplevels(tissue_gene_expression$y[ind])
x <- tissue_gene_expression$x[ind, ]
x <- x[, sample(ncol(x), 10)]
```

4\. In this case, LDA fits two 10-dimensional normal distributions. Look at the fitted model by looking at the `finalModel` component of the result of train. Notice there is a component called `means` that includes the estimate `means` of both distributions. Plot the mean vectors against each other and determine which predictors (genes) appear to be driving the algorithm. 

5\. Repeat exercises 3 with QDA. Does it have a higher accuracy than LDA?

6\. Are the same predictors (genes) driving the algorithm? Make a plot as in exercise 3.

7\. One thing we see in the previous plot is that the value of predictors correlate in both groups: some predictors are low in both groups while others are high in both groups. The mean value of each predictor, `colMeans(x)`, is not informative or useful for prediction and often, for interpretation purposes, it is useful to center or scale each column. This can be achieved with the `preProcessing` argument in `train`. Re-run LDA with `preProcessing = "scale"`. Note that accuracy does not change but see how it is easier to identify the predictors that differ more between groups in the plot made in exercise 4.

8\. In the previous exercises, we saw that both approaches worked well. Plot the predictor values for the two genes with the largest differences between the two groups in a scatterplot to see how they appear to follow a bivariate distribution as assumed by the LDA and QDA approaches. Color the points by the outcome.

9\. Now we are going to increase the complexity of the challenge slightly: we will consider all the tissue types.

```{r, eval = FALSE}
set.seed(1993)
y <- tissue_gene_expression$y
x <- tissue_gene_expression$x
x <- x[, sample(ncol(x), 10)]
```

What accuracy do you get with LDA?


10\. We see that the results are slightly worse. Use the `confusionMatrix` function to learn what type of errors we are making.


11\. Plot an image of the centers of the seven 10-dimensional normal distributions.

12\. Make a scatterplot along with the prediction from the best fitted model.


13\. Use the `rpart` function to fit a classification tree to the `tissue_gene_expression` dataset. Use the `train` function to estimate the accuracy. Try out `cp` values of `seq(0, 0.05, 0.01)`. Plot the accuracy to report the results of the best model.

14\. Study the confusion matrix for the best fitting classification tree. What do you observe happening for placenta?

15\. Notice that placentas are called endometrium more often than placenta. Note also that the number of placentas is just six, and that, by default, `rpart` requires 20 observations before splitting a node. Thus it is not possible with these parameters to have a node in which placentas are the majority. Rerun the above analysis, but this time permit `rpart` to split any node by using the argument `control = rpart.control(minsplit = 0)`. Does the accuracy increase? Look at the confusion matrix again.

16\. Plot the tree from the best fitting model obtained in exercise 11.

17\. We can see that with just six genes, we are able to predict the tissue type. Now let's see if we can do even better with a random forest. Use the `train` function and the `rf` method to train a random forest. Try out values of `mtry` ranging from, at least, `seq(50, 200, 25)`. What `mtry` value maximizes accuracy? To permit small `nodesize` to grow as we did with the classification trees, use the following argument: `nodesize = 1`. This will take several seconds to run. If you want to test it out, try using smaller values with `ntree`. Set the seed to 1990.

18\. Use the function `varImp` on the output of `train` and save it to an object called `imp`.

19\. The `rpart` model we ran above produced a tree that used just six predictors. Extracting the predictor names is not straightforward, but can be done. If the output of the call to train was `fit_rpart`, we can extract the names like this:

```{r, eval = FALSE}
ind <- !(fit_rpart$finalModel$frame$var == "<leaf>")
tree_terms <- 
  fit_rpart$finalModel$frame$var[ind] |>
  unique() |>
  as.character()
tree_terms
```

What is the variable importance in the random forest call for these predictors? Where do they rank?

20\. Extract the top 50 predictors based on importance, take a subset of `x` with just these predictors and apply the function `heatmap` to see how these genes behave across the tissues. We will introduce the `heatmap` function in @sec-clustering.



21\. FIX Previously, we have compared conditional probability give two predictors $p(x_1,x_2)$ to the fit $\hat{p}(x_1,x_2)$ obtained with a machine learning algorithm by making image plots. The following code can be used to make these images and include a curve at the values of $x_1$ and $x_2$ for which the function is $0.5$:

```{r}
#| eval: false
plot_cond_prob <- function(x_1, x_2, p){
  data.frame(x_1 = x_1, x_2 = x_2, p = p) |>
    ggplot(aes(x_1, x_2)) +
    geom_raster(aes(fill = p), show.legend = FALSE) +
    stat_contour(aes(z = p), breaks = 0.5, color = "black") +
    scale_fill_gradientn(colors = c("#F8766D", "white", "#00BFC4"))
}
```

We can see the true conditional probability for the 2 or 7 example like this:

```{r}
#| eval: false
with(mnist_27$true_p, plot_cond_prob(x_1, x_2, p))
```


Fit a kNN model and make this plot for the estimated conditional probability. Hint: Use the argument `newdata = mnist27$train` to obtain predictions for a grid points.

22\. Notice that, in the plot made in exercise 1, the boundary is somewhat wiggly. This is because kNN, like the basic bin smoother, does not use a kernel. To improve this we could try loess. By reading through the available models part of the **caret** manual, we see that we can use the `gamLoess` method. We need to install the __gam__ package, if we have not done so already. We see that we have two parameters to optimize:

```{r}
modelLookup("gamLoess")
```

Use cross-validation to pick a span between 0.15 and 0.75. Keep `degree = 1`. What span does cross-validation select?


23\. Show an image plot of the estimate $\hat{p}(x,y)$ resulting from the model fit in the exercise 2. How does the accuracy compare to that of kNN? Comment on the difference between the estimate obtained with kNN. 

24\. Use the `mnist_27` training set to build a model with several of the models available from the **caret** package. For example, you can try these:

```{r, eval = FALSE}
models <- c("glm", "lda",  "naive_bayes",  "svmLinear", "gamboost",  
            "gamLoess", "qda", "knn", "kknn", "loclda", "gam", "rf", 
            "ranger","wsrf", "Rborist", "avNNet", "mlp", "monmlp", "gbm", 
            "adaboost", "svmRadial", "svmRadialCost", "svmRadialSigma")
```

We have not explained many of these, but apply them anyway using `train` with all the default parameters. Keep the results in a list. You might need to install some packages. Keep in mind that you will likely get some warnings.

25\. Now that you have all the trained models in a list, use `sapply` or `map` to create a matrix of predictions for the test set. You should end up with a matrix with `length(mnist_27$test$y)` rows and `length(models)` columns.

26\. Compute accuracy for each model on the test set.

27\. Build an ensemble prediction by majority vote and compute its accuracy.

28\. Earlier we computed the accuracy of each method on the training set and noticed they varied. Which individual methods do better than the ensemble?

29\. It is tempting to remove the methods that do not perform well and re-do the ensemble. The problem with this approach is that we are using the test data to make a decision. However, we could use the accuracy estimates obtained from cross validation with the training data. Obtain these estimates and save them in an object.

30\. Now let's only consider the methods with an estimated accuracy of 0.8 when constructing the ensemble. What is the accuracy now?

31\. FIX If two methods give results that are the same, ensembling them will not change the results at all. For each pair of metrics compare the percent of time they call the same thing. Then use the `heatmap` function to visualize the results. Hint: use the `method = "binary"` argument in the `dist` function.

32\. Note that each method can also produce an estimated conditional probability. Instead of majority vote, we can take the average of these estimated conditional probabilities. For most methods, we can the use the `type = "prob"` in the train function. Note that some of the methods require you to use the argument `trControl=trainControl(classProbs=TRUE)` when calling train. Also, these methods do not work if classes have numbers as names. Hint: change the levels like this:

```{r, eval = FALSE}
dat$train$y <- recode_factor(dat$train$y, "2"="two", "7"="seven")
dat$test$y <- recode_factor(dat$test$y, "2"="two", "7"="seven")
```

33\. In this chapter, we illustrated a couple of machine learning algorithms on a subset of the MNIST dataset. Try fitting a model to the entire dataset.
