# Measurement error models

Another major application of linear models occurs in measurement errors models. In these situations, non-random covariates, such as time, are frequently encountered, with randomness often arising from measurement errors rather than from sampling or inherent natural variability.

## Example: modeling a falling object

To understand these models, imagine you are Galileo in the 16th century trying to describe the velocity of a falling object. An assistant climbs the Tower of Pisa and drops a ball, while several other assistants record the position at different times. Let's simulate some data using the equations we currently know and adding some measurement error. The **dslabs** function `rfalling_object` generates these simulations:

```{r, message=FALSE, warning=FALSE, cache=FALSE}
library(tidyverse)
library(broom)
library(dslabs)
falling_object <- rfalling_object()
```

The assistants hand the data to Galileo, and this is what he sees:

```{r gravity}
falling_object |> 
  ggplot(aes(time, observed_distance)) + 
  geom_point() +
  ylab("Distance in meters") + 
  xlab("Time in seconds")
```

Galileo does not know the exact equation, but by looking at the plot above, he deduces that the position should follow a parabola, which we can write like this:

$$
f(x) = \beta_0 + \beta_1 x + \beta_2 x^2
$$

The data does not fall exactly on a parabola. Galileo knows this is due to measurement error. His helpers make mistakes when measuring the distance. To account for this, he models the data with:

$$
Y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \varepsilon_i, i=1,\dots,n
$$

with $Y_i$ representing distance in meters, $x_i$ representing time in seconds, and $\varepsilon$ accounting for measurement error. The measurement error is assumed to be random, independent from each other, and having the same distribution for each $i$. We also assume that there is no bias, which means the expected value $\mbox{E}[\varepsilon] = 0$.

Note that this is a linear model because it is a linear combination of known quantities ($x$ and $x^2$ are known) and unknown parameters (the $\beta$s are unknown parameters to Galileo). Unlike our previous examples, here $x$ is a fixed quantity; we are not conditioning.

## Estimating parameters with least squares

To pose a new physical theory and start making predictions about other falling objects, Galileo needs actual numbers, rather than unknown parameters. Using LSE seems like a reasonable approach. How do we find the LSE?

LSE calculations do not require the errors to be approximately normal. The `lm` function will find the $\beta$s that will minimize the residual sum of squares:

```{r}
fit <- falling_object |> 
  mutate(time_sq = time^2) |> 
  lm(observed_distance~time+time_sq, data = _)
tidy(fit)
```

Let's check if the estimated parabola fits the data. The **broom** function `augment` allows us to do this easily:

```{r falling-object-fit}
augment(fit) |> 
  ggplot() +
  geom_point(aes(time, observed_distance)) + 
  geom_line(aes(time, .fitted), col = "blue")
```

Thanks to my high school physics teacher, I know that the equation for the trajectory of a falling object is:

$$
d(t) = h_0 + v_0 t -  0.5 \times 9.8 \, t^2
$$

with $h_0$ and $v_0$ the starting height and velocity, respectively. The data we simulated above followed this equation, adding measurement error to simulate `n` observations for dropping the ball $(v_0=0)$ from the tower of Pisa $(h_0=55.86)$.

These are consistent with the parameter estimates:

```{r}
tidy(fit, conf.int = TRUE)
```

The Tower of Pisa height is within the confidence interval for $\beta_0$, the initial velocity 0 is in the confidence interval for $\beta_1$ (note the p-value is larger than 0.05), and the acceleration constant is in a confidence interval for $-2 \times \beta_2$.


## Exercises 

1\. Plot CO2 levels for the first 12 months of the `co2` dataset and notice it seems to follow a sin wave with a frequency of 1 cycle per month. This means that a measurement error model that might work is 

$$
y_i = \mu + A \sin(2\pi \,t_i / 12 + \phi) + \varepsilon_i
$$
with $t_i$ the month number for observation $i$. Is this a linear model for the parameters $mu$, $A$ and $\phi$?


2\. Using trigonometry, we can show that we can rewrite this model as:

$$
y_i = \beta_0 + \beta_1 \sin(2\pi t_i/12) + \beta_2 \cos(2\pi t_i/12) + \varepsilon_i
$$

Is this a linear model? 

3\. Find least square estimates for the $\beta$s using `lm`. Show a plot of $y_i$ versus $t_i$ with a curve on the same plot showing $\hat{Y}_i$ versus $t_i$.

4\. Now fit a measurement error model to the entire `co2` dataset that includes a trend term that is a parabola as well as the sine wave model.

5\. Run diagnostic plots for the fitted model and describe the results.



